---
title: "Getting Started: Diamonds"
author: "Zachary del Rosario"
date: 2020-06-25
output:
  github_document
---

*Purpose*: Throughout this course, you'll complete a large number of *exercises* and *challenges*. Exercises are meant to introduce content with easy-to-solve problems, while challenges are meant to make you think more deeply about and apply the content. The challenges will start out highly-scaffolded, and become progressively open-ended.

In this challenge, you will go through the process of exploring, documenting, and sharing an analysis of a dataset. We will use these skills again and again in each challenge.

```{r setup}
library(tidyverse)
```

# Data Exploration
<!-- -------------------------------------------------- -->

In this first stage, you will explore the `diamonds` dataset and document your observations.

__q1__ Create a plot of `price` vs `carat` of the `diamonds` dataset below. Document your observations from the visual.

*Hint*: We learned how to do this in `e-vis00-basics`!

```{r q1-task}
## TASK: Plot `price` vs `carat` below
diamonds %>%
  ggplot(aes(carat, price)) +
  geom_point()
```

**Observations**:

- `price` generally increases with `carat`
- The trend is not 'clean'; there is no single curve in the relationship

__q2__ Create a visualization showing variables `carat`, `price`, and `cut` simultaneously. Experiment with which variable you assign to which aesthetic (`x`, `y`, etc.) to find an effective visual.

```{r q2-task}
## TASK: Plot `price`, `carat`, and `cut` below
diamonds %>%
  ggplot(aes(carat, price, color = cut)) +
  geom_point(alpha = 1 / 3) +
  geom_smooth(se = FALSE) +
  scale_x_log10() +
  scale_y_log10()
```

**Observations**:

- `price` generally increases with `carat`
- The `cut` helps explain the variation in price;
  - `Ideal` cut diamonds tend to be more expensive
  - `Fair` cut diamonds tend to be less expensive

