---
title: "Getting Started: Diamonds"
author: "Zach del Rosario"
date: 2020-06-11
output:
  github_document:
    toc: true
---

*Purpose*: Throughout this course, you'll complete a large number of *exercises* and *challenges*. Exercises are meant to introduce content with easy-to-solve problems, while challenges are meant to make you think more deeply about and apply the content. The challenges will start out highly-scaffolded, and become progressively open-ended.

In this challenge, you will go through the process of exploring, documenting, and sharing an analysis of a dataset. We will use these skills again and again in each challenge.

<!-- include-rubric -->

```{r setup}
library(tidyverse)
library(rzdr)
```

# Background
<!-- -------------------------------------------------- -->

Since I'm presenting, I'll add a bit of background on the data:

*Aside*: It's always a good idea to do a `glimpse` and `summary` first:

```{r glimpse}
diamonds %>% glimpse()
```

The glimpse shows us what variables are in the dataset, and a handful of values
just to get a sample of what we have.

```{r summary}
diamonds %>% summary()
```

The summary gives us a set of standard statistics for each of the variables.
This gives us a sense of "scale" for each variable.

## Pricing and the 4 C's
<!-- ------------------------- -->

Diamond pricing is traditionally based on the "Four C's": cut, color, clarity, and carat.
Three of those variables are *factors*: discrete variables with fixed *levels*.

[Diamond cuts](https://en.wikipedia.org/wiki/Diamond_cut) are fairly elaborate,
with many different styles. However the [American Gem
Society](https://en.wikipedia.org/wiki/American_Gem_Society) publishes standards
for its own cut grade ranking, which ranges from `Fair` to `Ideal.

```{r levels-cut}
diamonds %>%
  pull(cut) %>%
  levels()
```

- `Fair` is the lowest (least valuable) level
- `Ideal` is the highest (most valuable) level

```{r levels-color}
diamonds %>%
  pull(color) %>%
  levels()
```

[Diamond colors](https://en.wikipedia.org/wiki/Diamond_color) range from colorless `D` to more colorful `J`. Grades above `J`
exist with more apparent color.

```{r levels-clarity}
diamonds %>%
  pull(clarity) %>%
  levels()
```

[Diamond clarity](https://en.wikipedia.org/wiki/Diamond_clarity) refers to the lack of inclusions (internal defects) and blemishes (surface defects).

- `I1` refers to the *included* category, meaning undesirable inclusions are visible to the naked eye. This is the lowest level.
- `IF` refers to *internally flawless*, which is the highest level in the dataset.
- `FL` refers to *flawless*, which is rare and refers to a diamond with neither inclusions nor blemishes.

[Carat](https://en.wikipedia.org/wiki/Carat_(mass)) is a unit of mass often used
for gems. This is a continuous variable. Generally we would expect higher carat
to correspond to a higher sale price.

# Data Exploration
<!-- -------------------------------------------------- -->

In this first stage, you will explore the `diamonds` dataset and document your observations.

__q1__ Create a plot of `price` vs `carat` of the `diamonds` dataset below. Document your observations from the visual.

*Hint*: We learned how to do this in `e-vis00-basics`!

```{r q1-task}
## TASK: Plot `price` vs `carat` below
diamonds %>%
  ggplot(aes(carat, price)) +
  geom_point()
```

**Observations**:

- `price` generally increases with `carat`
- The trend is not 'clean'; there is no single curve in the relationship

__q2__ Create a visualization showing variables `carat`, `price`, and `cut` simultaneously. Experiment with which variable you assign to which aesthetic (`x`, `y`, etc.) to find an effective visual.

```{r q2-task}
## TASK: Plot `price`, `carat`, and `cut` below
diamonds %>%
  ggplot(aes(carat, price, color = cut)) +
  geom_point(alpha = 1 / 3) +
  scale_x_log10() +
  scale_y_log10() +
  theme_common()
```

**Observations**:

- `price` generally increases with `carat`
  - Furthermore the trend is roughly linear on a log-log scale; this implies a power-law relation between `price` and `carat`
- The `cut` helps explain the variation in price;
  - `Ideal` cut diamonds tend to be more expensive
  - `Fair` cut diamonds tend to be less expensive

**Elaboration**:

The figure above is rather busy; let's make a simpler visual with smoothed trends.

```{r elaboration}
diamonds %>%
  ggplot(aes(carat, price, color = cut)) +
  geom_smooth() +
  scale_x_log10() +
  scale_y_log10() +
  theme_common() +
  theme(legend.position = "bottom")
```

**Observations**:

- The relationships between `price` and `carat` is very nearly linear when plotted on a log-log scale!
  - Furthermore, the slope seems to be constant across different `cut` values.
- At low `carat`, the `Fair` diamonds appear to be the most expensive. Is this a real trend?
  - Note that the confidence band for `Fair` overlaps the other curves, so this is already not a robust conclusion.
- At very high `carat`, the price of `Fair` diamonds again become most expensive.

Let's follow up the low-carat diamonds:

```{r focus-low-carat}
diamonds %>%
  filter(carat <= 0.3) %>%

  ggplot(aes(carat, price)) +
  geom_hex(bins = 10) +

  scale_x_continuous(breaks = c(0.2, 0.25, 0.3)) +
  viridis::scale_fill_viridis() +
  facet_wrap(~cut) +
  theme_common() +
  theme(legend.position = "bottom")
```

**Observations**:

- The `Fair` observations are far more sparse than other `cut`s in the `carat <= 0.3` range.
- Based on observation, the bulk of `Fair` diamonds do not appear to be of especially high `price`.
  - However, there is an outlier in `price` at `carat` near `0.3`. This may be dragging the trend line up.
- Ultimately, the small sample size, edge effect, and the outlier can explain the odd phenomenon with `Fair` in the smoothed trend above. This casts doubt on the conclusions that small `Fair` diamonds are genuinely more expensive.




# Bonus Observations

We haven't gotten to this point in the exercise sequence yet, but it's usually a good idea to start with 1-dimensional investigations of the data when doing EDA. There's a very interesting pattern just within the distribution `carat` values.

```{r bonus1}
diamonds %>%
  ggplot(aes(carat)) +
  geom_histogram(bins = 120) +
  theme_common()
```

**Observations**:

- Even with a large number of bins there are very tall peaks. Let's zoom in a bit to inspect closer.

```{r bonus2}
wid <- 0.01
diamonds %>%
  filter(carat <= 1.3) %>%
  ggplot(aes(carat)) +
  geom_histogram(boundary = wid / 2, binwidth = wid) +
  scale_x_continuous(
    breaks = c(0.3, 0.4, 0.5, 0.7, 0.9, 1, 1.2)
  ) +
  theme_common()
```

**Observations**:

- There appear to be peaks in `carat` at particular `0.1` increments.
- These peaks tend to have a sharp drop to their left (lower values), with a slower dropoff to their right (higher values).
- Since these diamonds are meant for sale, these patterns are likely reflect the desires of the market: i.e. people desire diamonds at these specific carat values, and will buy diamonds slightly above (but not slightly below) those values.
